Nail: A practical interface generator for data formats

### 2014-04-30 paper/osdi14/design.tex
more edits
Acked-by: Julian Bangert <jbangert@acm.org>

\paragraph{Integers and Constraints}
\begin{figure}[tb]
\begin{tabular}{ll}
\toprule
\bf Syntax & \bf Semantics \\
\midrule
\texttt{int32} & 32-bit signed integer \\
\multirow{2}{*}{\texttt{uint4}}
  & 4-bit unsigned integer, \\
  & returned as an 8-bit value \\
\multirow{2}{*}{\texttt{uint8 | 1..3}}
  & 8-bit unsigned integer, \\
  & $1\leq x \leq 3$ \\
\multirow{2}{*}{\texttt{uint16 | [1,2]}}
  & unsigned 16-bit integer, \\
  & $x \in \{1,2\}$ \\
\multirow{2}{*}{\texttt{int32 | [1,5..255,512]}}
  & signed 32 bit integer, \\
  & $x \in \{ 1, 512 \} \vee 5\leq x\leq 255$ \\
\bottomrule
\end{tabular}
\caption{Example Nail grammars for integer values.}
\label{fig:range}
\end{figure}

@@

\paragraph{S.}

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parsers results. The
\emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.
For example, \texttt{many uint8} represents an array of 8-bit unsigned
integers, and \texttt{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
comma-separated lists of decimal numbers.

@@

\begin{figure}[tb]

\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
dns_header = {
  id uint16
  qr uint1
  opcode uint4
  aa uint1 
  tc uint1
  rd uint1
  ra uint1
  uint3 = 0
  rcode uint4
}
\end{verbatim} 
\end{minipage}
~
\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
struct dns_header {
  uint16_t id;
  uint8_t qr;
  uint8_t opcode;
  uint8_t aa;
  uint8_t tc;
  uint8_t rd;
  uint8_t ra;
  uint8_t rcode;
};
\end{verbatim} 
\end{minipage}

\caption{Nail grammar (left) and data model (right) for a part of the
grammar for DNS packets.  The \texttt{uint3 = 0} grammar represents
3 bits of padding (filled with zeroes).}
\label{fig:dns-struct}
\end{figure}

@@

\paragraph{Constants.}
In some cases, not all bytes in a structure actually contain information, such as magic numbers or
reserved fields. Those fields can be represented in Nail grammars by constant fields in structures.
Constant fields do not correspond to a field in the internal model, but they are validated during
parsing and generated during output.  Constants can either have integer values, such as
\texttt{uint3=0} or string values for text-based protocols, e.g. \texttt{many uint8 = ``Foo''}.

In some protocols, there might be many ways to represent the same constant field
and there is no semantic difference between the different syntactic representations.
Nail therefore allows repeated constants, such as \texttt{many (uint8=' ')}, which parses any number of space characters, or
\texttt{|| uint8 = 0x90 || uint16 = 0x1F0F}, which parses two of the many representations for X86
NOP instructions, which are used as padding between basic blocks in an executable.

@@

For example, \texttt{word= <uint8='"'; many int8 | 'a'..'z'; uint8='"'>} parses a quoted
lowercase word into an array.

\paragraph{References.}
To support recursive structures, we include a reference combinator \texttt{*}, which is
syntactically equivalent to directly including the parser, but in the internal model adds a level of
indirection to the value. This allows recursive structures to be parsed, for example, balanced
parentheses could be parsed as follows: \texttt{paren = < uint8 ="("; optional *paren ;uint8=")">}

@@

If two options the same string could be parser by two options, generated output for the latter
option is not necessarily understood identically by the parser. However, actual data formats are
usually not ambiguous in this sense. Figure~\ref{fig:choice} demonstrates a simple choice
combinator.

Nail also includes an \texttt{optional} combinator, which attempts to recognise a value, but succeeds
without consuming input when it cannot recognise that value. Syntactically, \texttt{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.

\begin{figure}[tb]
\begin{verbatim}
choose {
  A = uint8 | 1..8
  B = uint16 | ..256
}
\end{verbatim}
\caption{A simple choice combinator that parses either an 8-bit unsigned
integer with a value between 1 and 8 (option A), or a 16-bit unsigned
integer with a value of at most 256 (option B).}
\label{fig:choice}
\end{figure}

@@

Nail only provides one built-in combinator, \texttt{n_of} that uses dependency fields, which is used
for count fields. For example, DNS labels are encoded as Pascal-style strings - a length followed by
the value of the label. The Nail grammar for DNS packets in Figure~\ref{fig:dns-full} represents
this pattern as follows:

\begin{verbatim}
{ @length uint8 | 1..64
  label n_of @length uint8 }
\end{verbatim}

@@

In order to handle complicated input functions, Nail allows the programmer to
break out of Nail's generated code in a controlled fashion, by manipulating dependent fields and
streams. 
By default, Nail implements streams with in-memory buffers, but the programmer can override this implementation.
While parsing or generating a Nail grammar, there is always an implicit \emph{current stream} and
referred to as \texttt{\$current} in the Nail language.

@@

Transformations can handle a wide variety of patterns in data formats, including the following: 
\paragraph{Offsets.}
We provide a transformation for handling offset fields that is invoked as follows:
\texttt{\$fragment transform offset\_u32(\$file\_stream, \@offset)}. 
They are implemented by these two pseudocode functions:
\begin{verbatim}
int offset_u32_parse(NailArena *tmp,
 NailStream *out_str,
 NailStream *in_current,const uint32_t *off){
 out_str = suffix of in_current at offset off
}
int offset_u32_generate(NailArena *tmp,
  NailStream *in_fragment,
  NailStream *out_current,
  uint32_t *off){
   off = position of out_current
   append in_fragment to out_current
}
\end{verbatim}

>>>>

\paragraph{Integers and Constraints}
\begin{figure}[tb]
\begin{tabular}{@{}ll@{}}
\toprule
\bf Syntax & \bf Semantics \\
\midrule
\cc{int32} & 32-bit signed integer \\
\multirow{2}{*}{\cc{uint4}}
  & 4-bit unsigned integer, \\
  & returned as an 8-bit value \\
\multirow{2}{*}{\cc{uint8 | 1..3}}
  & 8-bit unsigned integer, \\
  & $1\leq x \leq 3$ \\
\multirow{2}{*}{\cc{uint16 | [1,2]}}
  & unsigned 16-bit integer, \\
  & $x \in \{1,2\}$ \\
\multirow{2}{*}{\cc{int32 | [1,5..255,512]}}
  & signed 32 bit integer, \\
  & $x \in \{ 1, 512 \} \vee 5\leq x\leq 255$ \\
\bottomrule
\end{tabular}
\caption{Example Nail grammars for integer values.}
\label{fig:range}
\end{figure}

@@

\paragraph{Repetition.}

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parsers results. The
\emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.
For example, \cc{many uint8} represents an array of 8-bit unsigned
integers, and \cc{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
comma-separated lists of decimal numbers.

@@

\begin{figure}[tb]
\smaller[0.5]
\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
dns_header = {
  id uint16
  qr uint1
  opcode uint4
  aa uint1 
  tc uint1
  rd uint1
  ra uint1
  uint3 = 0
  rcode uint4
}
\end{verbatim} 
\end{minipage}
~
\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
struct dns_header {
  uint16_t id;
  uint8_t qr;
  uint8_t opcode;
  uint8_t aa;
  uint8_t tc;
  uint8_t rd;
  uint8_t ra;
  uint8_t rcode;
};
\end{verbatim} 
\end{minipage}

\caption{Nail grammar (left) and data model (right) for a part of the
grammar for DNS packets.  The \cc{uint3 = 0} grammar represents
3 bits of padding (filled with zeroes).}
\label{fig:dns-struct}
\end{figure}

@@

\paragraph{Constants.}
In some cases, not all bytes in a structure actually contain information, such as magic numbers or
reserved fields. Those fields can be represented in Nail grammars by constant fields in structures.
Constant fields do not correspond to a field in the internal model, but they are validated during
parsing and generated during output.  Constants can either have integer values, such as
\cc{uint3 = 0} or string values for text-based protocols, e.g. \cc{many uint8 = ``Foo''}.

In some protocols, there might be many ways to represent the same constant field
and there is no semantic difference between the different syntactic representations.
Nail therefore allows repeated constants, such as \cc{many (uint8=' ')}, which parses any number of space characters, or
\cc{|| uint8 = 0x90 || uint16 = 0x1F0F}, which parses two of the many representations for X86
NOP instructions, which are used as padding between basic blocks in an executable.

@@

For example, \cc{word=<uint8='"'; many int8 | 'a'..'z'; uint8='"'>} parses a quoted
lowercase word into an array.

\paragraph{References.}
To support recursive structures, we include a reference combinator \cc{*}, which is
syntactically equivalent to directly including the parser, but in the internal model adds a level of
indirection to the value. This allows recursive structures to be parsed, for example, balanced
parentheses could be parsed as follows: \cc{paren = < uint8 ="("; optional *paren ;uint8=")">}

@@

If two options the same string could be parser by two options, generated output for the latter
option is not necessarily understood identically by the parser. However, actual data formats are
usually not ambiguous in this sense.
For example, Figure~\ref{fig:choice} demonstrates a simple choice
combinator.

\begin{figure}[h!]
\smaller[0.5]
\begin{verbatim}
  choose {
    A = uint8 | 1..8
    B = uint16 | ..256
  }
\end{verbatim}
\caption{A simple choice combinator that parses either an 8-bit unsigned
integer with a value between 1 and 8 (option A), or a 16-bit unsigned
integer with a value of at most 256 (option B).}
\label{fig:choice}
\end{figure}

Nail also includes an \cc{optional} combinator, which attempts to recognise a value, but succeeds
without consuming input when it cannot recognise that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.

@@

Nail only provides one built-in combinator, \cc{n_of} that uses dependency fields, which is used
for count fields. For example, DNS labels are encoded as Pascal-style strings: a length followed by
the value of the label. The Nail grammar for DNS labels can represent
this pattern as follows:

{
\smaller[0.5]
\begin{verbatim}
  {
    @length uint8 | 1..64
    label n_of @length uint8
  }
\end{verbatim}
}

@@

In order to handle complicated input functions, Nail allows the programmer to
break out of Nail's generated code in a controlled fashion, by manipulating dependent fields and
streams. 
By default, Nail implements streams with in-memory buffers, but the programmer can override this implementation.
While parsing or generating a Nail grammar, there is always an implicit \emph{current stream} and
referred to as \cc{\$current} in the Nail language.

@@

Transformations can handle a wide variety of patterns in data formats, including the following: 
\paragraph{Offsets.}
We provide a transformation for handling offset fields that is invoked as follows:
\cc{\$fragment transform offset\_u32(\$file\_stream, \@offset)}. 
They are implemented by two functions, as shown in Figure~\ref{fig:xform-sig}.

\begin{figure}[h]
\smaller[0.5]
\input{code/offset-xform}
\caption{Pseudocode for two functions that implement the offset transform.}
\label{fig:xform-sig}
\end{figure}
